Lexical frequency mediates the facilitation effect of cognateness in bilingual lexical acquisition

### 2023-04-28 manuscript.qmd
Add appendix for syllable frequencies and new revisions

format:
  apaquarto-docx: default
  apaquarto-pdf:
    pdf-engine: xelatex
    documentmode: man
    fontsize: 12pt
    keep-tex: false
    floatsintext: true
    blank-lines-above-title: 2
    header-includes:
      - \usepackage{tipa}
      - \usepackage[pinyin]{babel}
      - \usepackage{CJKutf8}
      - \usepackage{fontspec}
      - \babelprovide[script=CJK, language={Chinese Simplified}]{chinese-simplified}
      - \babelfont[chinese-simplified]{rm}[Ligatures = Common]{Noto Serif CJK SC}
      - \babelfont[chinese-simplified]{sf}[Ligatures = Common]{Noto Sans CJK SC}

    include-in-header:
      - text: '\usepackage{lineno}\linenumbers'

author-note:
  blank-lines-above-author-note: 3
  disclosures:
    conflict-of-interest: |
        The authors declare no conflicts of interest with regard to the funding source of this study.
    financial-support: |
        This research was supported by grants from the Spanish Ministerio de Ciencia, Innovación y Universidades (PGC2018-101831-B-I00 and PRE2019-088165), and the Catalan Government [ICREA (Catalan Institution for Research and Advanced Studies) Academia 2019 award]. Gonzalo Garcia-Castro was supported by a fellowship of the Spanish Ministerio de Ciencia, Innovación y Universidades (FPI 2019).
    gratitude: |
        We are grateful to Chiara Santolin, Alicia Franco-Martínez, Cristina Rodríguez-Prada, Ege E. Özer, and the rest of the Speech Acquisition and Perception research group for their helpful feedback. We thank Xavier Mayoral, Silvia Blanch, and Cristina Cuadrado for their technical support. We also thank Cristina Dominguez and Katia Pistrin for their efforts in recruiting infants. We also thank all families and infants who participated in the experiments.

@@

color_scheme_set(rep(clrs[1], 6))
options(ggplot2.ordinal.fill = clrs,
        ggplot2.ordinal.colour = clrs,
        ggplot2.discrete.fill = clrs,
        ggplot2.discrete.colour = clrs,
        ggplot2.continuous.fill = scale_color_gradient,
        ggplot2.continuous.colour = scale_color_gradient,
        knitr.graphics.error = FALSE) 
```

@@

The foundations of word learning are in place early in age. From six months of age, infants start directing their gaze to some objects when hearing their labels, according to both experimental data [@jusczyk1995infants; @tincoff1999beginnings; @bergelson2012months; @bergelson2015early] and parental vocabulary reports [@fenson2007macarthurbates; @samuelson2021precision]. During the last half of their second year, new word forms are acquired at an increasingly fast rate [@goldfield1990early; @fenson1994variability; @mayor2011statistical; @bloom2002how; @bergelson2020comprehension]. Such early stages of lexical acquisition are characterised by substantial variation across children reflected, for instance, on the variability of the number of words they know [i.e., vocabulary size, @fenson1994variability; @frank2021variability]. Despite this variability, trajectories of early vocabulary growth seem quite stable across languages. @tardif2008baby collected data about the first ten words acquired by 10-- to 16--month-old infants living in the United States, Hong Kong, and Beijing. Since birth, these infants had been learning English, Cantonese, and Mandarin, respectively. The authors found a common pattern across the three groups: their first ten words referred to roughly the same concepts, namely relatives/caregivers (*daddy*, *mommy*), social routines (*bye*, *uh-oh*) or animal sounds (*woof-woof*). These results were later extended by @frank2021variability to a more diverse set of languages, and found that such cross-language commonalities are stronger at earlier stages of lexical acquisition.

Most research on early word acquisition relies extensively on data from monolingual children, and is oblivious to substantial proportion of the world population acquiring more than one language since early ages [@grosjean2021extent]. Previous research on bilingual vocabulary acquisition pointed to bilingual toddlers knowing, on average, less words in each of their languages than their monolinguals peers, and also to both groups knowing a similar amount of words when the bilinguals' two languages are taken into account. @hoff2012dual found that English-Spanish bilingual toddlers in South Florida (United States) knew less words in English than monolinguals did. Both groups knew a similar total amount of words when both English and Spanish vocabularies were counted together, which substantiates the importance of collecting data on both languages when assessing bilinguals' communicative development. Other studies have provided converging evidence that bilinguals know a similar or even larger number of words than monolinguals when the two languages are aggregated [@oller2002language; @pearson1994patterns; @pearson1993lexical; @patterson2004comparing; @patterson2004bilingual; @smithson2014bilingualism; @petitto2001bilingual; @gonzalez-barrero2020bilingual]. These studies rely on bilingual samples whose languages are relatively distant in typology: English is a Germanic language, while Spanish is a Romance language, and both languages differ substantially at the grammatical or the phonological level. Little is known about vocabulary acquisition in bilinguals learning a pair of typologically closer languages (e.g., when both are Romance languages).

For a given set of concepts, bilingual children may be exposed to two distinct sets of word-forms: one in each language. Depending on the similarity between both languages, the two sets of words may overlap in form in varying degrees. When both languages are typologically close, like Spanish and Catalan (both Romance languages), they are more likely to share a large amount of cognates (i.e., form-similar translation equivalents) than two typologically more distant languages, like Spanish and English (Romance and Germanic languages, respectively). For instance, in the presence of a door, a Spanish-Italian (or a Spanish-Catalan) bilingual might be exposed to the words *puerta* and *porta* (cognates), whereas a Spanish-English bilingual might hear *puerta* and *door* (non-cognates). It could be the case that mapping two phonologically similar labels (cognates like *puerta*-*porta*) onto the same referent is easier than doing the same with two phonologically dissimilar labels (non-cognates, like *puerta* and *door*). If cognates are easier to acquire than non-cognates, bilinguals learning two languages that share a high proportion of cognates should benefit more often from this facilitation effect than bilinguals learning two languages with a lower proportion of cognates, and should therefore show larger vocabulary sizes.

@floccia2018introduction provided evidence in line with this claim. The authors collected vocabulary data on word comprehension and production from 372 24-month-old bilingual toddlers living in the United Kingdom who were learning English and an additional language. The additional language was one of 13 typologically diverse languages: Bengali, Cantonese Chinese, Dutch, French, German, Greek, Hindi/Urdu, Italian, Mandarin Chinese, Polish, Portuguese, Spanish and Welsh. The authors calculated the average phonological overlap between the words in each of these additional languages and their translation equivalents in English, which was taken as a proxy of the degree of *cognateness* between each pair of languages. Phonological overlap was measured by computing the Levenshtein distance between each cross-language pair of phonetic transcriptions. The Levenshtein distance is a metric that computes the distance between two strings of characters/phonemes by counting the smallest number of insertions, deletions and substitutions one of the strings has to undergo to become identical to the other [@levenshtein1966binary]. For instance, the Levenshtein distance for the translation equivalent *cat*-*kat* (/kæt/-/\textipa{kAt}/) in English and Dutch would be one: a single edition would be needed to make both phonological translation identical (replacing /æ/ by /\textipa{A}/). The same translation equivalent in English and Greek (*cat*-*$\gamma\acute{\alpha}\tau\alpha$*, /kæt/-/\textipa{Gata}/) would score three, as it requires at least three editions: substituting /k/ for /\textipa{G}/, substituting /æ/ for /a/ and adding another /a/.

Levenshtein distance scores are divided by the length of the longest string to normalise the similarity scores between 0 and 1. Finally, the result is subtracted from one, to get the complementary, and therefore a measure of phonological similarity, instad of dissimilarity. Following the previous examples: $1/4 = 0.25$ for /kæt/-/\textipa{kAt}/, and $3/4 = 0.75$ for /kæt/-/\textipa{Gata}/. Finally, this variable was included as a predictor of participants' vocabulary sizes. Among other findings, Floccia and co-workers reported an increase in vocabulary size in the additional language (i.e., not English) associated with an increase in the average phonological similarity between the translation equivalents of each language pair. For example, English-Dutch bilinguals (0.2214 mean phonological overlap), were able to produce more Dutch words than English-Mandarin bilinguals (0.0197 mean phonological overlap) were able to produce in Mandarin.

Floccia et al. pointed to *parallel activation* as the underpinning mechanism behind their results. The parallel activation hypothesis stems from the language non-selective account of lexical access, which suggests that bilinguals activate both languages simultaneously during speech production [@costa2000cognate; @hoshino2008cognate] or comprehension [@spivey1999cross; @thierry2007brain], even in fully monolingual environments. One of the clearest pieces of evidence of parallel activation was provided by @costa2000cognate. In this study, Spanish monolinguals and Catalan-Spanish bilingual adults were asked to name pictures of common objects in Spanish. In half of the trials, the object labels were cognates in Spanish and Catalan (*árbol*-*arbre*, translations of *tree*), whereas in the other half of the trial labels were non-cognates (*mesa*-*taula*, translations of *table*), importantly such distinction was only relevant for bilinguals. Bilinguals named cognate pictures faster than non-cognate pictures, even after adjusting for the lexical frequency of the items. In contrast, Spanish monolinguals, who were unfamiliar with the Catalan translations of the Spanish words they uttered, showed equivalent naming times for the two types of stimuli. These results showed that bilinguals' Catalan phonology was activated during the production of Spanish words, facilitating the naming of cognate pictures.

Regarding word acquisition, the parallel activation account is a plausible explanation for Floccia et al.'s results in the light of two observations. First, bilingual toddlers' lexicon seems to consist of a large proportion of translation equivalents: the presence of a word in their lexicon is a good predictor of its translation equivalent also being present [@bilson2015semantic; @tsui2022are; @pearson1995crosslanguage]. Second, there is experimental evidence of parallel activation during word comprehension in children and toddlers [@vonholzen2012language; @poulin-dubois2013lexical; @bosma2020cognate]. It is possible that cognateness increases the amount of cross-language activation, facilitating word acquisition. This could ultimately lead to children who are learning languages with a larger proportion of cognates (e.g., English and Dutch) showing larger vocabulary sizes. This account in which cognateness underlies the facilitation effect of typological language distance is also in line with previous studies suggesting that the acquisition of new words is facilitated by their phonological or semantic similarity with other words [@hills2009longitudinal; @fourtassi2020growth; @laing2022phonological; @jones2019children]. If phonological similarity plays a cross-language facilitation role during word acquisition, cognate translation equivalents---which share both semantic and phonological similarity---should be acquired, on average, earlier that non-cognate translation equivalents---which share semantic similarity (e.g., are taxonomically related) but not phonological similarity.

@@

More recently, @mitchell2022cognates addressed this issue providing a more fine grained measure of a  cognateness facilitation effect in vocabulary acquisition. Using a longitudinal sample of 47 16-to-30 month-old French-English bilinguals living in Canada, the authors collected data on expressive vocabulary data in both languages. They created two lists of translation equivalents: one made of 131 cognates, and one made of 406 non-cognates. The proportion of translation equivalents that children were reportedly able to produce was higher in the cognate lists than in the non-cognate list across ages, even when both lists were matched by semantic category (furniture, animals, food were similarity represented in both lists) and age of acquisition norms (an index of word difficulty). These findings shed some light on the ongoing exploration of why (if at all) bilinguals' vocabulary size seems to grow faster when both languages are more similar at the lexical level. Word-forms sharing more phonemes with their translation equivalents (i.e., cognates) seem to be acquired faster. Parallel activation remains a plausible explanation for this effect. When a child is exposed to a word form, its (phonologically similar) translation equivalent is activated, and this cross-language activation increases the chances of acquisition. The precise underpinnings of this effect are still unclear.

@@

This facilitation would occur as the result of the semantic and phonological overlap between both words. When the child has sufficient lexical knowledge of *gat*, and therefore there is some for-referent association in place, future presentations of the similar sounding translation *gato* in the presence of the same reference should lead to an increased likelihood of acquisition. In the case of a non-cognate translation equivalent, such facilitation would be unlikely. For instance, even if the child has already acquired the word *gos*, hearing its dissimilar Spanish translation *perro* would hardly lead to an increased probability of acquisition. In line with this account, only the later-acquired form of the translation equivalent would benefit from its cognate status. Given that children are more likely to acquire words from languages to which they are more often exposed to [@david2008individual; @cattani2014how; @thordardottir2006bilingual], the acquisition of words in the language of lower exposure should, on average, be more susceptible to the effect of cognateness. For instance, if the previous child is exposed 80% of the time to English, and 20% of the time to Spanish, words in Spanish (especially low-frequency ones) would show a cognateness facilitation effect, with an earlier age of acquisition for cognates than for non-cognates.

@@

We tested the hypothesis that the cognate effect on word acquisition is conditional to a child's exposure rate to the word, which would support the hypothesis that cognateness plays a role in the acquisition of a word *only* once its translation has been already acquired. We collected vocabulary data on production and comprehension from a large sample of bilingual Catalan-Spanish children using an online vocabulary checklist designed for the present study. Participants were aged `r min(floor(participants$age))` to `r max(floor(participants$age))` months, and were exposed to Catalan and/or Spanish to varying degrees. We adopted a Bayesian explanatory item response theory  [IRT, see @kachergis2022standard for a similar approach] approach to model the probability of a given participant being reported to *understand* or *understand and say* each word-form in the checklist, conditional to participants age and rate of exposure to the word-form, and the cognate status of the word form, while adjusting for the the number of phonemes in the word form (an indicator of word acquisition difficulty).

@@

@@

```{r apatb-loos}
#| label: apatb-results-loos
#| echo: false
#| message: false
#| warning: false
#| eval: true
#| apa-cap: "Bayesian leave-one-out cross validation (LOO-CV)."
#| apa-note: "We report the ELPD of each model along their associated standard errors (SE). The last column indicates the difference in ELPD between each model and the model with the best predictive accuracy (Model 3). Pareto-k estimates of all models were acceptable (k < 0.5)."
model_loos |>
    loo_compare() |>
    as.data.frame() |>
    rownames_to_column("model") |>
    mutate(model = paste0("Model ", str_remove(model, "model_fit_"))) |>
    select(model, elpd_loo, se_elpd_loo, elpd_diff, se_diff) |> 
    gt(rowname_col = "model") |>
    fmt_number(2:5, decimals = 2, sep_mark = ",") |>
    cols_merge_uncert(elpd_loo, se_elpd_loo) |> 
    cols_merge_uncert(elpd_diff, se_diff) |> 
    cols_label(
        model = "Model",
        elpd_loo = "ELPD $$\\pm$$ SE",
        elpd_diff = "Diff. $$\\pm$$ SE"
    ) |>
    tab_style(cell_text(weight = "bold", style = "italic"),
              cells_column_labels()) |>
    tab_style(cell_text(align = "left"),
              cells_title(groups = "subtitle"))
```

@@

posterior_draws$summary |>
    select(.variable_name, .median, .lower, .upper, .rope_overlap, .type) |>
    mutate(across(.median:.upper,
                  \(x) ifelse(.type==intercept_label, plogis(x), x/4))) |> 
    gt(groupname_col = ".type", 
       rowname_col = ".variable_name") |>
    fmt_percent(.rope_overlap) |> 
    fmt_number(c(.median:.upper)) |>
    cols_merge(c(.lower, .upper),
               pattern = "[{1}, {2}]") |>
    cols_label(
        .variable_name = "Parameter",
        .median = "Median",
        .lower = md("95% HDI"),
        .rope_overlap = md("ROPE prob.")
    ) |>
    tab_style(cell_text(weight = "bold"),
              cells_column_labels(columns = 1:5)) |>
    tab_style(cell_text(align = "left"),
              cells_title(groups = "subtitle")) |> 
    tab_style(cell_text(size = "small"), 
              list(cells_body(),
                   cells_row_groups(),
                   cells_stub())) |> 
    tab_style(cell_text(style = "italic"),
              cells_row_groups())
```

@@

```{r apafg-results-marginal}
#| label: apafg-results-marginal
#| echo: false
#| message: false
#| warning: false
#| fig-width: 8
#| fig-height: 6
#| eval: true
#| apa-cap: "Posterior marginal effects of Model 3. Thin lines correspond to 50 mean predictions generated from 50 MCMC samples of the posterior distribution of the model. Thicker lines indicate the median of those predictions. Different colours indicate different levels of phonological similarity (as indicated by the Cognateness similarity between pairs of translation equivalents). Predictions are presented separately for different degrees of word exposure index: little exposure to the word, mean exposure, and high exposure. Predictions for Comprehension are show on top and predictions for Comprehension and Production are shown on the bottom. In-sample predictions lie inside the grey rectangles."

@@

{{< include appendix/_model-predictors.qmd >}}

@@

{{< include appendix/_model-details.qmd >}}

@@

{{< include appendix/_model-diagnostics.qmd >}}

@@

{{< include appendix/_model-doe.qmd >}}

>>>>

format:
  # apaquarto-docx: default
  apaquarto-pdf:
    pdf-engine: xelatex
    documentmode: man
    fontsize: 12pt
    keep-tex: false
    floatsintext: true
    blank-lines-above-title: 2
    header-includes:
      - \usepackage{tipa}
    include-in-header:
      - text: '\usepackage{lineno}\linenumbers'

author-note:
  blank-lines-above-author-note: 3
  disclosures:
    conflict-of-interest: |
        The authors declare no conflicts of interest with regard to the funding source of this study.
    financial-support: |
        This research was supported by grants from the Spanish Ministerio de Ciencia, Innovación y Universidades (PGC2018-101831-B-I00 and PRE2019-088165), and the Catalan Government [ICREA (Catalan Institution for Research and Advanced Studies) Academia 2019 award]. Gonzalo Garcia-Castro was supported by a fellowship of the Spanish Ministerio de Ciencia, Innovación y Universidades (FPI 2019).
    gratitude: |
        We are grateful to Chiara Santolin, Ege E. Özer, Alicia Franco-Martínez, Cristina Rodríguez-Prada, and the rest of the Speech Acquisition and Perception research group for their helpful feedback. We thank Xavier Mayoral, Silvia Blanch, and Cristina Cuadrado for their technical support, and Cristina Dominguez and Katia Pistrin for their efforts in recruiting infants. We also thank all families and infants who participated in the experiments.

@@

color_scheme_set(rep(clrs[1], 6))

options(ggplot2.ordinal.fill = clrs,
        ggplot2.ordinal.colour = clrs,
        ggplot2.discrete.fill = clrs,
        ggplot2.discrete.colour = clrs,
        ggplot2.continuous.fill = scale_color_gradient,
        ggplot2.continuous.colour = scale_color_gradient,
        knitr.graphics.error = FALSE) 
```

@@

The foundations of word learning are in place early in age. From six months of age, infants start directing their gaze to some objects when hearing their labels, according to both experimental data [@jusczyk1995infants; @tincoff1999beginnings; @bergelson2012months; @bergelson2015early] and parental vocabulary reports [e.g., @fenson2007macarthurbates; @samuelson2021precision]. During the last half of the second year of life, new word-forms are acquired at an increasingly fast rate [@goldfield1990early; @fenson1994variability; @mayor2011statistical; @bloom2002how; @bergelson2020comprehension]. Such early stages of lexical acquisition are characterised by substantial heterogeneity across different populations of language learners, reflected for instance on (the variability of) quantitative indices of lexical development, like vocabulary size [i.e., vocabulary size, @fenson1994variability; @frank2021variability]. Despite this variability, trajectories of early vocabulary growth seem quite stable across languages. @tardif2008baby collected data about the first ten words acquired by 10-- to 16--month-old infants living in the United States, Hong Kong, and Beijing. Since birth, these infants had been learning English, Cantonese, and Mandarin, respectively. The authors found a common pattern across the three groups: their first ten words referred to roughly the same concepts, namely relatives/caregivers (*daddy*, *mommy*), social routines (*bye*, *uh-oh*) or animal sounds (*woof-woof*). These results were later extended by @frank2021variability to a more diverse set of languages, and found that such cross-language commonalities are stronger at earlier stages of lexical acquisition.

Most research on early word acquisition relies extensively on data from monolingual children, and is oblivious to the fact that a substantial proportion of the world population acquires more than one language from early ages [@grosjean2021extent]. Previous research on bilingual vocabulary acquisition pointed to bilingual toddlers knowing, on average, less words in each of their languages than their monolinguals peers, and also to both groups knowing a similar amount of words when the bilinguals' two languages are taken into account. @hoff2012dual found that English-Spanish bilingual toddlers in South Florida (United States) knew less words in English than monolinguals did. Both groups knew a similar total amount of words when both English and Spanish vocabularies were counted together, which substantiates the importance of collecting data on both languages when assessing bilinguals' communicative development. Other studies have provided converging evidence that bilinguals know a similar or even larger number of words than monolinguals when the two languages are aggregated [@oller2002language; @pearson1994patterns; @pearson1993lexical; @patterson2004comparing; @patterson2004bilingual; @smithson2014bilingualism; @petitto2001bilingual; @gonzalez-barrero2020bilingual]. These studies rely on bilingual samples whose languages are relatively distant in typology: English is a Germanic language, while Spanish is a Romance language, and both languages differ substantially at the grammatical or the phonological level. Little is known about vocabulary acquisition in bilinguals learning a pair of typologically closer languages (e.g., when both are Romance languages).

For a given set of concepts, bilingual children may be exposed to two distinct sets of word-forms: one in each language. These cross-language word pairs that share meaning are referred to as *translation equivalents*. Throughout the present study, we make the assumption that bilinguals form common semantic representations for both languages (e.g., the concept of a *chair* is shared for the English and Spanish words 'chair' and 'silla')^[The referential contexts in which bilingual children hear (and learn) both labels may not be identical. For instance, the English word *finger* refers to eight parts of the body (excluding thumbs and toes), whereas its Spanish translation *dedos* refers to twenty body parts (fingers, thumbs, and toes). These discrepancies are reflected, for instance, in adults' performance in lexical decision tasks [@chaouch-orozco2023are]. The fact that some translation pairs share more semantic features than others does not, however, inform whether bilinguals form common or separate semantic representations for translation equivalents. What English speakers understand for 'fingers' might not be identical to what Spanish speakers understand for 'dedos'. But when an English speaker learns Spanish, it seems implausible that they would form a separate semantic representation for what 'dedo' conveys at the conceptual level, compared to the possibility that they associate the new label to the semantic representation they already had formed for 'fingers'. Therefore, the lack of complete semantic alignment is not necessarily an argument against the existence of common semantic representations for translation equivalents. For simplicity, and in line with previous (formal and verbal) models of bilingual lexical processing [e.g., @shook2013bilingual; @dijkstra2019multilink; @dijkstra1998bia; @kroll2010revised], we assume common semantic representations for translation pairs.]. Depending on the similarity between both languages, the two sets of words may overlap in form in varying degrees. When both languages are typologically close, like Spanish and Catalan (both Romance languages), they are more likely to share a large amount of cognates (i.e., form-similar translation equivalents) than two typologically more distant languages, like Spanish and English (Romance and Germanic languages, respectively). For instance, in the presence of a door, a Spanish-Italian (or a Spanish-Catalan) bilingual might be exposed to the words *puerta* and *porta* (cognates), whereas a Spanish-English bilingual might hear *puerta* and *door* (non-cognates). It could be the case that mapping two phonologically similar labels (cognates like *puerta*-*porta*) onto the same referent is easier than doing the same with two phonologically dissimilar labels (non-cognates, like *puerta* and *door*). If cognates are easier to acquire than non-cognates, bilinguals learning two languages that share a high proportion of cognates should benefit more often from this facilitation effect than bilinguals learning two languages with a lower proportion of cognates, and should therefore show larger vocabulary sizes.

@floccia2018introduction provided evidence for such a facilitation effect of typological distance and vocabulary size. The authors collected vocabulary data on word comprehension and production from 372 24-month-old bilingual toddlers living in the United Kingdom who were learning English and an additional language. The additional language was one of 13 typologically diverse languages: Bengali, Cantonese Chinese, Dutch, French, German, Greek, Hindi/Urdu, Italian, Mandarin Chinese, Polish, Portuguese, Spanish and Welsh. The authors calculated the average phonological overlap between the words in each of these additional languages and their translation equivalents in English, which was taken as a proxy of the degree of *cognateness* between each pair of languages. Phonological overlap was measured by computing the Levenshtein distance between each cross-language pair of phonetic transcriptions. The Levenshtein distance is a metric that computes the distance between two strings of characters/phonemes by counting the smallest number of insertions, deletions and substitutions one of the strings has to undergo to become identical to the other [@levenshtein1966binary]. For instance, the Levenshtein distance for the translation equivalent *cat*-*kat* (/kæt/-/\textipa{kAt}/) in English and Dutch would be one: a single edition would be needed to make both phonological translation identical (replacing /æ/ by /\textipa{A}/). The same translation equivalent in English and Greek (*cat*-*$\gamma\acute{\alpha}\tau\alpha$*, /kæt/-/\textipa{Gata}/) would score three, as it requires at least three editions: substituting /k/ for /\textipa{G}/, substituting /æ/ for /a/ and adding another /a/. Levenshtein distance scores are divided by the length of the longest string to normalise the similarity scores between 0 and 1. Finally, the result is subtracted from one, to get the complementary, and therefore a measure of phonological similarity, instead of dissimilarity. Following the previous examples: $1/4 = 0.25$ for /kæt/-/\textipa{kAt}/, and $3/4 = 0.75$ for /kæt/-/\textipa{Gata}/. Finally, this variable was included as a predictor of participants' vocabulary sizes. Among other findings, Floccia and co-workers reported an increase in vocabulary size in the additional language (i.e., not English) associated with an increase in the average phonological similarity between the translation equivalents of each language pair. For example, English-Dutch bilinguals (0.2214 mean phonological overlap), were able to produce more Dutch words than English-Mandarin bilinguals (0.0197 mean phonological overlap) were able to produce in Mandarin. More recently, @blom2020crosslanguage and @bosma2019longitudinal reported similar results, providing converging evidence of for facilitatory effect of language distance on vocabulary size estimates.

Floccia et al. pointed to *parallel activation* as the underpinning mechanism behind their results. The parallel activation hypothesis stems from the language non-selective account of lexical access, which suggests that bilinguals activate both languages simultaneously during speech production or comprehension, even in fully monolingual environments. Evidence supporting the language-non selective account of lexical access is not restricted to production or the auditory modality, but rather extend to comprehension and translation, and to the visual modality (reading and signing) [@spivey1999cross; @hoshino2008cognate; @schwartz2007reading; @schwartz2006bilingual; @morford2011when; @vanhell2008sentence; @gimeno-martinez2021crosslinguistic; see @kroll2013multilingual for review]. One of the clearest pieces of evidence of parallel activation was provided by @costa2000cognate. In this study, Spanish monolinguals and Catalan-Spanish bilingual adults were asked to name pictures of common objects in Spanish. In half of the trials, the object labels were cognates in Spanish and Catalan (*árbol*-*arbre*, translations of *tree*), whereas in the other half of the trial labels were non-cognates (*mesa*-*taula*, translations of *table*), importantly such distinction was only relevant for bilinguals. Bilinguals named cognate pictures faster than non-cognate pictures, even after adjusting for the lexical frequency of the items. In contrast, Spanish monolinguals, who were unfamiliar with the Catalan translations of the Spanish words they uttered, showed equivalent naming times for the two types of stimuli. These results showed that bilinguals' Catalan phonology was activated during the production of Spanish words, facilitating the naming of cognate pictures.

Bilingual toddlers and children---still in the process to developing a mature lexicon---also show evidence of parallel activation during comprehension [@vonholzen2012language; @vonholzen2019impact; @bosma2020cognate; @poarch2012crosslanguage; @schroter2016orthographic; @poulin-dubois2013lexical]. Whether---and if so, how---the language non-selectivity shapes the early bilingual lexicon is still unclear. Previous work has pointed to an interactive account of lexical acquisition, in which toddlers' lexical representations in one language influence the acquisition of words in the other language. This cross-language facilitation is, in principle, possible since early ages: the bilingual infant lexicon consists in a substantial proportion of  translation equivalents: the presence of a word in the lexicon is a good predictor of its translation equivalent also being present [@bilson2015semantic; @tsui2022are; @pearson1995crosslanguage; @pearson1995crosslanguage]. In fact, translation equivalents may even be overrepresented in the early bilingual lexicon. @bilson2015semantic found that bilingual toddlers produced a higher proportion of translation equivalents than it would be expected if they learned two independent lexica simultaneously, which suggests that bilingual children preferentially acquired words for which they already knew a label in the other language. In the same line, @bosch2014first reported a similar overrepresentation of translation equivalents in Catalan-Spanish bilinguals: a high proportion of acquired Catalan words held an also acquired translation in Spanish. Finally, a recent investigation by @tsui2022area presented a quantitative model of bilingual vocabulary acquisition which showed a consistent preferential acquisition of translation equivalents by English-French bilinguals, with such preference being stronger at early stages of vocabulary growth. Overall, these findings suggest that bilingual children benefit from a semantic facilitation effect driven by the acquisition of translation equivalents.

It is possible that cognateness increases the amount of cross-language activation, facilitating early word acquisition. This could ultimately lead to children who are learning languages with a larger proportion of cognates (e.g., English and Dutch) showing larger vocabulary sizes, as found by Floccia et al. There is evidence of for phonological and semantic similarity facilitating word acquisition in monolinguals: infants are more likely to acquired new words that are semantically or phonologically more connected to other words, than words with poorer connectivity in either of those dimensions, such as words belonging to sparse phonological neighbourhoods [@hills2009longitudinal; @fourtassi2020growth; @laing2022phonological; @jones2019children; @storkel2004children]. If phonological similarity plays a cross-language facilitation role during bilingual word acquisition, cognate translation equivalents---which share both semantic and phonological similarity---should be acquired, on average, earlier that non-cognate translation equivalents---which share semantic similarity (e.g., are taxonomically related) but not phonological similarity.

@@

More recently, @mitchell2022cognates addressed this issue providing a more fine grained measure of a  cognateness facilitation effect in vocabulary acquisition. Using a longitudinal sample of 47 16-to-30 month-old French-English bilinguals living in Canada, the authors collected data on expressive vocabulary data in both languages. They created two lists of translation equivalents: one made of 131 cognates, and one made of 406 non-cognates. The proportion of translation equivalents that children were reportedly able to produce was higher in the cognate lists than in the non-cognate list across ages, even when both lists were matched by semantic category (furniture, animals, food were similarity represented in both lists) and age of acquisition norms (an index of word difficulty). These findings shed some light on the ongoing exploration of why (if at all) bilinguals' vocabulary size seems to grow faster when both languages are more similar at the lexical level. Word-forms sharing more phonemes with their translation equivalents (i.e., cognates) seem to be acquired faster. Parallel activation remains a plausible explanation for this effect. When a child is exposed to a word-form, its (phonologically similar) translation equivalent is activated, and this cross-language activation increases the chances of acquisition. The precise underpinnings of this effect are still unclear.

@@

This facilitation would occur as the result of the semantic and phonological overlap between both words. When the child has sufficient lexical knowledge of *gat*, and therefore there is some for-referent association in place, future presentations of the similar sounding translation *gato* in the presence of the same reference should lead to an increased likelihood of acquisition. In the case of a non-cognate translation equivalent, such facilitation would be unlikely. For instance, even if the child has already acquired the word *gos*, hearing its dissimilar Spanish translation *perro* would hardly lead to an increased probability of acquisition. In line with this account, only the later-acquired form of the translation equivalent would benefit from its cognate status. Given that children are more likely to acquire words from languages to which they are more often exposed to [@david2008individual; @cattani2014how; @thordardottir2006bilingual; @hoff2012dual], the acquisition of words in the language of lower exposure should, on average, be more susceptible to the effect of cognateness. For instance, if the previous child is exposed 80% of the time to English, and 20% of the time to Spanish, words in Spanish (especially low-frequency ones) would show a cognateness facilitation effect, with an earlier age of acquisition for cognates than for non-cognates.

@@

We tested the hypothesis that the cognate effect on word acquisition is conditional to a child's exposure rate to the word, which would support the hypothesis that cognateness plays a role in the acquisition of a word *only* once its translation has been already acquired. We collected vocabulary data on production and comprehension from a large sample of bilingual Catalan-Spanish children using an online vocabulary checklist designed for the present study. Participants were aged `r min(floor(participants$age))` to `r max(floor(participants$age))` months, and were exposed to Catalan and/or Spanish to varying degrees. We adopted a Bayesian explanatory item response theory  [IRT, see @kachergis2022standard for a similar approach] approach to model the probability of a given participant being reported to *understand* or *understand and say* each word-form in the checklist, conditional to participants age and rate of exposure to the word-form, and the cognate status of the word-form, while adjusting for the the number of phonemes in the word-form (an indicator of word acquisition difficulty).

@@

Finally, we included word length as a nuance predictor in our analyses, motivated by previous studies reporting a strong association between word length and vocabulary size estimates [e.g., @braginsky2019consistency; @jones2019children]. Sorter word-forms like /\textipa{sO\textltilde}/ (Catalan for *sun*) are generally simpler to encode as phonological representations in the lexicon than longer word-forms like /\textipa{ku.ku"DRi\textltilde}/ (Catalan for *crocodile*), and are therefore more likely to be participant a child's vocabulary.

@@

```{r apatb-loos}
#| label: apatb-results-loos
#| echo: false
#| message: false
#| warning: false
#| eval: true
#| apa-cap: "Bayesian leave-one-out cross validation (LOO-CV)."
#| apa-note: "We report the ELPD of each model along their associated standard errors (SE). The last column indicates the difference in ELPD between each model and the model with the best predictive accuracy (Model 3). Pareto-k estimates of all models were acceptable (k < 0.5)."
model_loos |>
    loo_compare() |>
    as.data.frame() |>
    rownames_to_column("model") |>
    mutate(model = paste0("Model ", str_remove(model, "model_fit_"))) |>
    select(model, elpd_loo, se_elpd_loo, elpd_diff, se_diff) |> 
    gt(rowname_col = "model") |>
    fmt_number(2:5, decimals = 2, sep_mark = ",") |>
    cols_merge_uncert(elpd_loo, se_elpd_loo) |> 
    cols_merge_uncert(elpd_diff, se_diff) |> 
    cols_label(
        model = "Model",
        elpd_loo = "ELPD (SE)",
        elpd_diff = "ELPD difference (SE)"
    ) |>
    tab_style(cell_text(weight = "bold", style = "italic"),
              cells_column_labels()) |>
    tab_style(cell_text(align = "left"),
              cells_title(groups = "subtitle"))
```

@@

posterior_draws$summary |>
    select(.variable_name, .median, .lower, .upper, .rope_overlap, .type) |>
    mutate(across(.median:.upper,
                  \(x) ifelse(.type==intercept_label, plogis(x), x/4))) |> 
    gt(groupname_col = ".type", 
       rowname_col = ".variable_name") |>
    fmt_percent(.rope_overlap, decimals = 3) |> 
    fmt_number(c(.median:.upper), decimals = 3) |>
    cols_merge(c(.lower, .upper),
               pattern = "[{1}, {2}]") |>
    cols_label(
        .variable_name = "Parameter",
        .median = "Median",
        .lower = md("95% HDI"),
        .rope_overlap = md("ROPE prob.")
    ) |>
    tab_style(cell_text(weight = "bold"),
              cells_column_labels(columns = 1:5)) |>
    tab_style(cell_text(align = "left"),
              cells_title(groups = "subtitle")) |> 
    tab_style(cell_text(size = "small"), 
              list(cells_body(),
                   cells_row_groups(),
                   cells_stub())) |> 
    tab_style(cell_text(style = "italic"),
              cells_row_groups())
```

@@

```{r apafg-results-marginal}
#| label: apafg-results-marginal
#| echo: false
#| message: false
#| warning: false
#| fig-width: 8
#| fig-height: 6
#| out-width: 80%
#| eval: true
#| apa-cap: "Posterior marginal effects of Model 3. Thin lines correspond to 50 mean predictions generated from 50 MCMC samples of the posterior distribution of the model. Thicker lines indicate the median of those predictions. Different colours indicate different levels of phonological similarity (as indicated by the Cognateness similarity between pairs of translation equivalents). Predictions are presented separately for different degrees of word exposure index: little exposure to the word, mean exposure, and high exposure. Predictions for Comprehension are show on top and predictions for Comprehension and Production are shown on the bottom. In-sample predictions lie inside the grey rectangles."

@@

{{< include _appendix/_model-predictors.qmd >}}

@@

{{< include _appendix/_model-details.qmd >}}

@@

{{< include _appendix/_model-diagnostics.qmd >}}

@@

{{< include _appendix/_model-doe.qmd >}}
